# Data Structures and Types {#data-struc}

There is no scalar in R. 

There is no scalar in R. 

There is no scalar in R. 

Every structure is of *vector* form. A (seemingly) scalar is indeed a vector of length 1. This is a striking concept compared to other general-purposed language or statistical packages. It turns out that the fact won't affect mush about how to code R, but the understanding of such concept is helpful for better coding thinking in the R language.

Compared to other general-purposed language, the built-in data structures are not rich in R. But they are considerably flexible. Most of the analytics can be fulfilled based on the minimal set of these structures.

All structures are vectors in R. And there are two types of vector: atomic and recursive. This chapter will quickly go through both of them.

## Atomic Vector

> Common types of atomic vector include: numeric, character, logical, and factor.

Atomic vector is the foundamental data structure in R. It is called atomic because it can only contain the same type of data and no nesting allowed. For example:

```{r}
1
```

The number `1` is a vector (of length 1, so it looks like a scalar but yes it is a vector). To check the type of a vector one can use the function `typeof`:

```{r}
typeof(1)
```

The result of `typeof` reveals that the storage mode of the vector `1` is double; hence the vector is a numeric vector. There are many bulit-in functions to manipulate numeric vector. Some examples follow:

```{r}
1:5
seq(5, 1, -1)
c(1, 1, 2, 3, 5)
rep(777, 3)
```

One can also use `integer`, `numeric`, `character`, and `logical` as a creation function to create vectors of corresponding type, with the argument to be the desired length of the created vector:

```{r}
integer(3)    # default value is 0L
numeric(3)    # default value is 0
character(3)  # default value is "" (empty string)
logical(3)    # default value is FALSE
```

Normally a number will be associated with type `double`. To force R use `integer` type (it use considerably less storage space compared to a `double`), just append `L` to the number:

```{r}
typeof(1:5L)
```

Since an atomic vector is, atomic, it has no nesting structure:

```{r}
c(1:5, seq(5, 1, -1), c(1, 1, 2, 3, 5))
```

The **result of concatenation of multiple atomic vectors is one single atomic vector**, not a vector nested with three different vectors. Does this mean that atomic vectors of different types can not be combined? Not necessary. 

```{r}
num_vec <- 1:3
str_vec <- c('a', 'b', 'c')
c(num_vec, str_vec)
```

When a numeric vector and a character vector combine, the resulting atomic vector is of type `character`. Such behavior is called type **coercion** and is a very important concept in most programming languages. Coercion always occurs when different types of atomic vectors are combined. The principle of coercion is to minimize information loss, if any.

For example, when a logical vector combined with a numeric:

```{r}
num_vec <- 1:3 # here one uses "<-"" to asign a value to an object
bol_vec <- c(TRUE, FALSE, FALSE) # or c(T, F, F) but not recommended
typeof(c(num_vec, bol_vec))
```

logicals are coerced into `integer`. 

More examples on coercion:

```{r}
typeof(c(1, 2L))
typeof(c(1L, TRUE))
typeof(c(1.0, TRUE))
typeof(c(1.0, TRUE, "foo"))
```

In addition to coercion, one can also explicitly convert one type to another. This is done by the function family `as`. For example, to convert logicals into integers and vice versa:

```{r}
as.logical(-1:2L)
as.integer(c(TRUE, FALSE, FALSE))
```

Notice that when casting integers to logicals, all the non-zeroes are converted to `TRUE`, only 0s are converted to `FALSE`.

### General Operations on Vectors

#### Subseting
Vectors can be subset by using the bracket syntax.

```{r}
vv <- 10:1
vv[1:2]
vv[c(1, 3, 5)]
vv[rep(3, 5)]
```

The bracket accepts *another vector as a selection vector*. If the selection vector is of type `numeric`, the original vector is to be selected based on numerical index. Any non-integer will be `floor`ed so the result of `c(1:10)[c(1.1, 1.9)]` is `r c(1:10)[c(1.1, 1.9)]`.

Since the bracket accepts vectors, one can also use logical vector for subsetting:

```{r}
vv <- 3:1
vv[c(TRUE, FALSE, FALSE)]
```

The fact that subsetting is done by another vector results in filtering being very easy:

```{r}
vv <- 1:10
vv[vv > 5]
```

Since all structures are vectors, mathematical operators such as `>` also work on  vectors by nature. The result of `vv > 5` is an element-wise comparison and hence also a vector of the same lenght:

```{r}
vv <- 1:10
vv > 5
```

The resulting logical vector can then be used to subset the original vector.

One can also use negative selection:

```{r}
vv <- 1:3
vv[-3]
vv[-length(vv)]
vv[-(1:3)]
```

Subset a vector may resulting in missing values:

```{r}
vv <- 1:3
vv[4]
```

Here the selection vector tries to extract the 4th element of the original vector but there is none. This will not cause any error in R; instead, the resulting vector will contains `NA`, which generally means a missing value.

#### Updating
Vectors can be updated by using the bracket syntax as well.

```{r}
vv <- 1:5
vv[1:3] <- 0
vv
```

One interesting question arises: what happen if one tried to update an element that does not exist in the vector?

```{r}
vv <- 1:5
vv[7] <- 0
vv
```

Two things to be noticed. First, the value will be asigned, as if it is newly created. Second, the vector will be expanded with `NA`s to align the new length that satisfies the updated result.

How about a delete? There is no delete method in R. To effectively delete an element in a vector, use negative selection and asign the new object:

```{r}
# to delete the 5th element:
vv <- 1:5
(vv <- vv[-5])  # use parenthesis to force print
```

The operation of the bracket syntax is indeed *functional*. (Type `?"["` to see the document.) To understand more about what's going on behibd the scene, one should refer to section \@ref(extraction-function) and also \@ref(replacement-function).

#### Naming
Vectors can be named. To check the names of a vector, use the function `names`. To name or rename a vector, just try asign the names in a character vector to the `names` function call.

```{r}
vv <- 1:3
names(vv)
names(vv) <- c("foo", "bar", "baz")
str(vv)
```

If one is confused why the renaming syntax actually works, see section \@ref(replacement-function) for more details.

Vectors can be partially named:

```{r}
vv <- 1:3
names(vv) <- c("foo", "bar")
names(vv)
```

Since the third element is unnamed, it is `NA` when calling `names` to print the names.

### Factor
Factors are special integer vectors. They are generally used to record categorical variables. It combines both features of characters and numerics, so it could be confusing for new users. Usually a factor is defined on a set of characters:

```{r}
baz <- c("foo", "bar")
bazf <- factor(baz)
typeof(bazf)
str(bazf)
```

Notice that the `typeof` indicates an `integer` type. This is true because factors are stored internally as integers. The `levels` of a factor reveals the complete set of characters that could appear in the factor. Now consider a little more complicated example:

```{r}
baz <- c("foo", "bar", "bar", "foo", "bar")
bazf <- factor(baz)
levels(bazf)
```

When an object contains repeating samples of characters, it may be a good choice to use `factor` type to represent it for performance and efficiency issue, since `levels` only contain distinct values.

`levels` can be manually specified when creating a `factor`, and `levels` are ordered. The order will be automatically determined if not specified. 

```{r}
baz <- c("foo", "bar", "bar", "foo", "bar")
bazf1 <- factor(baz, levels=c("bar", "foo"))
bazf2 <- factor(baz, levels=c("foo", "bar"))
str(bazf1)
str(bazf2)
```

`levels` can even contain unseen values:

```{r}
baz <- c("foo", "bar", "bar", "foo", "bar")
bazf3 <- factor(baz, levels=c("foo", "bar", "baz"))
str(bazf3)
```

### Matrix

### Array

## Recursive Vector

### List

### Data Frame


## Type Conversion

## Special Values
### `NA`
### `NaN`
### `NULL`


