# String Operations {#string}

## Basics
### `paste`
Strings are called `character` in the R language. It's important to know some basic string manipulation skills. One useful string function is the `paste` function:

```{r}
paste("foo", "bar", "baz", sep=' ')
```

The function takes as many arguments as it can and concatenates them with the separator specified in the `sep` arguemnt. Notice that `character` is vector, so the `paste` function can operate on vector--this is called vectorization in R:

```{r}
paste(c("foo", "bar"), "baz1" , sep=' ')
paste(c("foo", "bar"), c("baz1", "baz2", "baz3") , sep=' ')
```

What happens here is that the shorter arguments are *recycled* to have equal length and then each string is concatenated elment-wise. For more details about vectorization and recycling, see section \@ref(char).

The `paste` function can also *collapse* a character vector into one character (a vector of length 1). This is done by using the `collapse` argument:

```{r}
paste(c("foo", "bar", "baz"), collapse='-')
```

Collapsing can also be used *after* concatenation. This is done by supplying more than one character vectors to `paste` and also specifying the `collapse` argument:

```{r}
# concatenate only
paste(c("foo", "bar", "baz"), letters[1:5])
# concatenate and collapse
paste(c("foo", "bar", "baz"), letters[1:5], collapse='-')
```

There is also a variant of `paste` called `paste0`. The only difference is that `paste0` use a default separator of empty string (i.e., use `sep=''` in `paste`):

```{r}
paste0(c("foo", "bar"), "baz")
```

### `sprintf`
Another useful function is `sprintf`. It is used to format a *string with interpolation*:

```{r}
sprintf("hello %s!", "world")
sprintf("hello %.3f!", 7123)
sprintf("hello %i", 7)
```

See `?sprintf` for more formatting options. Multiple interpolation is also supported, given in order:

```{r}
sprintf("first: %i; second: %s", 1, "two")
```

### `cat` and `print`
The `cat` function concatenates its arguments and print.

```{r}
cat("foo", c("bar", "baz"))
cat("foo", c("bar", "baz"), sep=':')
```

Notice the difference between `cat` and `paste`: the former does not recycle and the result is printed; the latter does recylcing and the result is *returned* rather than printed. Too see this fact clear:

```{r}
# letters is a built-in variable storing all lower-cased English alphabets
# .Last.value is another built-in variable storing the result of the last evaluation
cat(letters[1:3], sep='+')
.Last.value
paste(letters[1:3], collapse='+')
.Last.value
```

### Miscellaneous

Use `substr` to extract partial string. Use `nchar` to count number of characters in a string. Use `strsplit` to split a string by specified separator.

```{r}
ss <- "why so serious?"
substr(ss, 1, 3)
nchar(ss)
strsplit(ss, split=' ') # the returned type is a list
```

## Regular Expressions
There are some built-in functions in the R language for using regular expressions. Some of them are quite useful, say, `gsub` and `grep`. But some are not that intuitive to work with.

### Filtering
Use `grep` to match and return the matched. It is vectorized, can return either numeric index or original charaters by using the `value` argument. One can also choose to return a `logical` by using the `grepl` variant.

### Substitution
Use `gsub` for substitution of matched pattern.

### Extraction
There are multiple ways to do matched pattern extraction using built-in functions. Sadly none of them are intuitive to use. The most useful technique may be the combination of `regexec` and `regmatches`.

```{r}
s <- "$VAR/foo/bar"
matched <- regexec("\\$([^/]*)", s)
regmatches(s, matched)
```

However, it is still too much to type. The coding style can be simplified by using the `magrittr` library:

```{r}
library(magrittr)
"$VAR/foo/bar" %>% regmatches(., regexec("\\$([^/]*)", .)) %>% .[[1]]
```

### The `stringr` Library
The `stringr` provides more intuitive and easy-to-use functions regarding regex operations.

Four particularly ussful functions are:

+ `str_extract` and `str_extract_all`
+ `str_match` and `str_match_all`
