# Control Flow {#control}

## Conditioning

### `if`

The `if` statement (a function indeed) can handle conditional branching of the codes. It can be followed by optional `else if` and `else` branches.

```{r}
if ( TRUE ) {
  print("Yes it is true.")
} else if ( TRUE ) {
  print("This is never printed.")
} else {
  print("When nothing above holds, this is printed.")
}
```

Notice that when `else if` exists, **conditions are checked in order**. The first matched branch will be executed and all the other dsicarded. This is why the second branch is never triggered in the above example even if the condition is `TRUE` as well.

The condition in `if` can not handle logical vectors with length more than one. In the following example, a warning will be issued with the result clearly explained.

```{r, warning=TRUE}
if ( c(TRUE, FALSE) ) {
  1 + 1
}
```

This is why sometimes people think there does exist scalar variable in R. But the truth is that everything is in vector form.

There is a more functional way of using `if`. Consider this example:

```{r}
# set.seed(777)
cond <- sample(1:10, 1)
res <- if ( cond > 5 ) "foo" else "bar"
res
```

The above code is stochastic due to `sample` without a fixed `seed`. Here one directly asign the result of an `if` to a variable. Such syntax only works in functional language. It works because `if` is a function, and a function should always return something--here evaluation of the last expression of `if` is the returned value of it.

When the conditioning is more complicated, a recommended indent style may look like:

```{r}
# set.seed(777)
cond <- sample(1:10, 1)
res <- 
  if ( cond > 5 ) {
    message("cond is ", cond, ": go if")
    "foo" 
  } else {
    message("cond is ", cond, ": go else")
    "bar"
  }
res
```

### `ifelse`

There is a vectorized version of `if`: the `ifelse` function. Now it accepts a vector of any length:

```{r}
vec_cond <- sample(1:10, 10, replace=TRUE) > 5
ifelse(sample(1:10, 10, replace=TRUE) > 5, 1:10, -(1:10))
```

Notice that the call to `sample` results in a vector of length 10, the second argument is a vector when the condition holds, and the third for condition not holds. What happen if the vector lengths are different? *Recycling* will occur. See section \@ref(recycling) for more details about recycling.

### Logical Operations

There are in general two types of logical operators in R. The single form (like `&`, `|`) and the double form (like `&&` annd `||`). The former is vectorized, i.e., element-wise operator, and the latter is scalar-like: only the first element is processed. To see things clear:

```{r}
c(TRUE, FALSE, FALSE) & -1:1 
c(TRUE, FALSE, FALSE) && -1:1 
```

Users who are familiar with other programming language may feel confused because in most general-purposed language only the double form is used as logical operator and the single form is used to perform bit-wise operation. In the R language this is simply not true. The single form serves both as bit-wise and also vectorized logical operator. Use `xor` for vectorized logical AND bit-wise XOR operator.

One should also note that there is implicit casting (or coercion) happening in the above example. Numeric vectors are coerced into logical ones before a logical operation actually takes place.

## Loop

### `repeat`

To repeat a code chunk, use `repeat`. Since there is no conditioning in `repeat`, usually users must also specify a conditonal `break` by `if` to avoid an infinite loop. 

```{r}
cnt <- 0
repeat {
  print("Hello World!")
  cnt <- cnt + 1
  if ( cnt >= 5 )
    break
}
```

### `while`

The `while` loop is another common control flow available in most programming language. It accepts a condition and as long as the condition holds the code chunk will repeat. The condition is checked at the very begining of each run.

```{r}
cnt <- 0
while ( cnt < 5 ) {
  print("Hello while!")
  cnt <- cnt + 1
}
```

There is no "until" loop in R. One can use `repeat` to implement an "until" function. Also a `while ( TRUE )` implementation can be readily replaced with `repeat.` 

### `for`

Another must-have loop structure is the `for` loop. It implements iteration operations. The basic syntax of `for` would look like:

```{r}
for ( i in 1:5 )
  print(i)
```

Remember that one can use `if` functionally. That is, we cab asign the returned value of a `if` to a variable. Is that also true for `for`? Consider the following codes:

```{r}
res <- for ( i in 1:10 ) i
res
i
```

Oops! the variable `res` has a value of `NULL`, not a value of the last evaluated `i` (which shall be 10). Does this mean that `for` is not a function? No. Every operation in R is a function call. `if` is a function. `for` is a function. But `for` is a function that will return `NULL` by default and that is not alterable. More interesting, `for` actually returns `NULL` *invisibly*, so one won't get a `NULL` printed to the console everytime when a `for` is done.

Iterating over numbers is not all what `for` can do. Basically a `for` can iterate on any vector, atomic or recursive. This is powerful and is often ignored by beginners.

```{r}
a <- rnorm(100)
b <- runif(100)
for ( l in list(a,b) )
  print(mean(l))
```

All vectors can be iterated. What about a `matrix`? Remember that a `matrix` is simply an atomic `vector` with a `dim` attribute. So surely it can be iterated: (only for illustration purpose because such scenario is rarely practical)

```{r}
for ( i in matrix(1:4,2,2) ) print(i)
```

`for` is useful. But don't use `for` for large-sized iteration. The overhead of `for` is usually quite high and since most basic operations in R are vectorzied, one should consider vectorizing in the first place when developing scripts. The following somewhat silly coding illustrates the basic difference in performance when vectorization is available:

```{r}
tt <- vv <- kk <- integer(1e5)
system.time(for ( i in 1:length(tt) ) vv[i] <- tt[i] + 1)
system.time(kk <- tt + 1)
identical(vv, kk)
```

To understand more, see the section \@ref(vectorization).


