# Function {#func}

Functions are very important in the R language because R is mostly a functional language. Indeed, **every operation ever happens is a function call**. This may not be easily seen in the first place. But while one types something like:

```{r}
1 + 1
```

what really happens is a function call to the add function (see `?"+"`). The expression is a sugar: this is the usual way human beings write down a mathematical expression. The functional way of such expression would be more like this one:

```{r}
`+`(1, 1)
```

This fact holds not only for `+`, but also other operators like `-`, `*`, `/`, `>`, `<`, and `=`. Indeed these so-called "operators" are just functions. Control flows are also functions (see `?"if"`). Indeed, even `{` and `[` are functions.

Functions are objects in R. So they can be manipulated in the way objects can be. Let's do some impractical but potentially inspiring coding:

```{r}
for ( op in c(`+`, `-`, `*`, `/`) )
  print(op(10, 2))
```

Functions are objects. And objects can be asigned. Here one asigns a function to another variable and use that variable to call the function. Another question in hand: is ``c(`+`, `-`, `*`, `/`)`` an atomic vector or recursive one? There is no `function`-typed aotomic vector. So even if the elements are all `function` (indeed, `builtin`), the resulting concatenation is a `list`.

## User-Define Function

A function can be created using the function `function`.

```{r}
f <- function() {}
```

In the aboce one liner a function `f` that does nothing is defined. A function can be called when suffixed with parantheiss `()`:

```{r}
f()
```

Notice that a `NULL` is return even if the function definition didn't do it explicitly. Functions should always return something. `NULL` is that something, for acutally nothing. To explicitly return a value in a function definition, one can use the `return` function:

```{r}
f <- function() {
  return(1)
}
f()
```

It is important to understand that a `return` is an exit point for a function. It signals the function to end its task. So consider the following function definition:

```{r}
f <- function() {
  print(0)
  return(1)
  print(2)
}
f()
```

The number 2 will never be printed because there is a `return` before it. One don't need a `return` to return values in a function. When there is no explicit `return`, a function will return its last-evaluated value:

```{r}
f <- function() {
  1 + 1
  2 + 2
}
f()
```

So why bothering use the `return`? It can be handy if one expects a function to exit in an early point by some conditioning:

```{r}
f <- function() {
  if ( sample(c(TRUE, FALSE), 1) ) {
    print("No Luck.")
    return(NULL)    
  }
  print("Do things...")
  1 + 1
  2 + 2
}
for ( i in 1:5 ) { f() }
```

The above function may randomly exit. 

## Arguemnts

A function can have arguemnts. Returned values are output of a function, arguments are input. Both are optional in coding but jusy the former will be implicitly set. To use arguments in a function, just name it in the definition:

```{r}
f <- function(x, y) {
  x
}
identical(f(1), f(2, x=1))
```

Several things worth noting:

+ There can be many arguments
+ Arguments are not typed: just as all other variables in R
+ Arguments can bve supplied by position or by name
+ Arguments that are not used are not checked at all (so can be missing)

```{r}
f <- function(x, y) {
  args <- as.list(match.call())
  for ( a in args ) print(a)
}
f(1, 2)
```

### Default Value

Arguments can have default value given in the definition. When actually supplied in the function call, the default value will be overwritten.

```{r}
f <- function(x, y=1) {
  x + y
}
for ( i in 1:5 ) print(f(i))
for ( i in 1:5 ) print(f(10, i))
```

### Ellipsis (`...`)


## Anonymous Function (Lambda)
In all the above examples, functions are asigned to variables. Or naturally speaking, functions are named. This is, however, not necessary. Consider the following example:

```{r}
f <- function(g) {
  typeof(g)
}
f(function() 1 + 1)
```

What's going on here? The function `f` is a function that print the result of `typeof` for its argument. Yes it is completely redundant and is purely created for educational purpose. The argument `g` can be virtaully anything. So why not a function? When one types things such as `f(3)` one doesn't asign a variable the number 3 (i.e., `x <- 3; f(x)`) so why bothering assigning a function? That's where the anonymous function comes in place. 

Sometimes a function is a one-timer so it is not worth tpying and assigning a variable for it. This is especailly true when using the `apply` family, which is explored in section \@ref(the-apply-family). Anonymous functions are sometimes called *lambda* in other programming language and is a common utility in modern languages, even for languages that is not functional (e.g., Python).

There is another use case when one won't explicitly assign variable to a function, because that will be done in another function! That is, one can write a function that returns a function:

```{r}
f <- function(x) {
  if ( x == "add" ) {
    function(y, z)
      y + z
  } else if ( x == "mul" ) {
    function(y, z)
      y * z
  } else {
    function()
      NULL
  }
}
g <- f("add") # the result should be 3
identical(g(1, 2), f("add")(1, 2))
```

Now the world is more functional. The `f` function is a factory to create different functions by its arguemnt. The return value of `f` is a newly created function--whatever name it is--and since the return value is a function, of course it can be called to work. Even the return function doesn't need to have be named or assigned, one can directly call it with `()` suffix.

## Lexical Scoping

Function definitions can be nested.

```{r}
f <- function(x) {
  g <- function(x) x + 1 # { is ignored if the body is one-lined
  g(x)
}
f(10)
```

## Extraction Function
Everything that happens is a function call. So things like these should be no exception.

```{r}
vv <- 1:10
vv[1:3]
```

The `<-` is the assignment function, i.e., `"<-"("vv", 1:10)` is what really happens in the functional style. And the second line, which is more interesting, is an extraction function of the form `[<-` taking place. (Type `?"[<-"` to see the docuemnt to confirm it realy is it.) So the second line is equivalent to:

```{r}
`[`(vv, 1:3)
```

That is not cool. Nobody writes in that (somewhat ugly?) way. So the R language provides the syntactic sugar to let users write codes in more human-readble fasion.

## Replacement Function
Even more interesting is the replacement function:

```{r}
vv <- 1:10
vv[1:3] <- 0
vv
```

The second line really has stories. Now one knows that `[` is indeed a function. An extraction function. So how come the second line works? Because one is now assigning a value to the return value of a function. Is that okay? Consider the following example:

```{r, error=TRUE}
f <- function(x) x
f(1) <- 2
```

No. One CAN'T assign a value to a function call. The reason why `vv[1:3] <- 0` will work is because it is completely another function call: the call to the replacement function `[<-`:

```{r}
vv <- 1:10
`[<-`(vv, 1:3, 0)
```



