# Function {#func}

Functions are very important in the R language because R is mostly a functional language. Indeed, **every operation ever happens is a function call**. This may not be easily seen in the first place. But while one types something like:

```{r}
1 + 1
```

what really happens is a function call to the add function (see `?"+"`). The expression is a sugar: this is the usual way human beings write down a mathematical expression. The functional way of such expression would be more like this one:

```{r}
`+`(1, 1)
```

This fact holds not only for `+`, but also other operators like `-`, `*`, `/`, `>`, `<`, and `=`. Indeed these so-called "operators" are just functions. Control flows are also functions (see `?"if"`). Indeed, even `{` and `[` are functions.

## User-Define Function

A function can be created using the function `function`.

```{r}
f <- function() {}
```

A function `f` that do nothing is defined.





## Anonymous Function (Lambda)

## Lexical Scoping

## Extraction Function

## Replacement Function

## Ellipsis (`...`)




