# R Language Characteristics {#char}

## Help
Use `?` to see the document of a function. One can even use `?` to see the document of `?`: try `?"?"`. The document, however, sometimes does not help a lot. The target audience of R document is usually not for beginners so may contain examples or wordings that are hard to understand. If one is consulting third-party packages not included in the core R, the quality of documents are indeed varying. Some authors provides good documentation while the others doesn't. That said, documents are only one source of help for R learners. Usually a better way is by googling and searching on [stackoverflow](http://stackoverflow.com/).

## Vectorization
Every built-in data structure in R is of vector form. This means that many of the built-in functions that operate on these data structures actually operates on vectors. Consider the following minimal example:

```{r}
1:10 + 1
```

Yes the add (`+`) function is vectorized. An expereicend R user won't write code like this:

```{r}
res <- integer(10)
for ( i in 1:10 )
  res[i] <- i + 1
res
```

The performance of vectorized codes and non-vectorized one can be huge:

```{r}
sillyAdd <- function(n) {
  res <- integer(n)
  for ( i in 1:n ) res[i] <- i + 1
  res
}
system.time(res1 <- 1:1e6 + 1)
system.time(res2 <- sillyAdd(1e6))
identical(res1, res2)
```

Due to performance (in most cases) gain, one should always consider using vectorization whenever possible.

One thing usually mis-understood by beginners: Are the `apply` family vectorized? The anwser is disppointedly a NO. `apply` family is a functional way (without side-effect) to write down codes intending to do loop-like tasks. They are NOT vectorization of any kind. This also means they do NOT gain any performance advantage.

## Recycling
Recycling usually happens when an operation involves in at least two vectors, and one is shorter than the other. Take extraction by logical vector for example:

```{r}
vec <- 1:10
vec[c(T, F)]
```

The logical vector `c(T, F)` is recycled to make sure it is as long as the vector to be extracted. Recycling often occurs without awaring users. No warning. No message. Recyclingis everywhere:

```{r}
1:10 + 1
```

Here the shorter vector (which is `1`) is recycled to have length 10. To see things more clear:

```{r}
identical(1:10 + 1:2, 1:10 + rep(1:2, 5))
```

## The `apply` Family
The `apply` family is really one outstanding feature in the core of R language. Among the family, the most useful two are `apply` and `lapply`. Sometimes `sapply` also comes in handy.

### `apply`
`apply` will call the given function column-by-column or row-by-row, according to the `MARGIN` argument supplied.

```{r}
(mm <- matrix(1:12, 4, 3))
apply(mm, 1, sum) # by the 1st dimension: row
apply(mm, 2, sum) # by the 2nd dimension: column
```

`apply` use ellipsis, so it is very easy to apply functions with additional arguments:

```{r}
mm <- matrix(1:12, 4, 3)
apply(mm, 2, mean, trim=.2) # apply the trimmed mean over each column
apply(mm, 2, paste, collapse=':') # concatenate all numbers for each column
```

or lambdas:

```{r}
mm <- matrix(1:12, 4, 3)
apply(mm, 2, function(x) sd(x) / mean(x))
```

`apply` can also be used on `data.frame` in exactly the same way:

```{r}
str(cars)
apply(cars, 2, max)
```

### `lapply` and `sapply`
`lapply` is a list `apply`, so it operates on a list.

```{r}
ll <- list(1:10, letters)
lapply(ll, length)
```

`lapply` always returns a list of the same length as its input. Sometimes it is sufficient to return only an atomic vector. Thgat's where `sapply` comes in handy:

```{r}
sapply(ll, length)
```

Most of the time the call to `sapply` will have the effect of a `unlist(lapply(...))`.


