# Working Directory {#work-env}

Use `getwd` to query the current working directory. To change working directory, use `setwd`. For Windows user, the path separator is typically the back-slash `\`. But it is the escape character in R so if one would like to use back-slash it must be escaped, for example, like this:

```r
setwd("C:\\Users\\kylechung")
```

or equivalently one could still follow *nix convention:

```r
setwd("C:/Users/kylechung")
```

One can use the function `dir` to return all the files and folders under the current directory:

```{r}
# the content of this book's working directory
dir()
```

`dir` can also have an argument `path` pointing to whereever one would like to examine the file contents. The default value of `path` is `.`, which is the notation indicating current directory. One can also use relative path notation: `..` to detnote the parent directory of current directory.

## Library Search Path

Use `.libPaths` to show the path(es) that will be searched by R.

```{r}
.libPaths()
```

## Save / Load

Use `save` to save serialized `.RData` for one variable. Or use `save.image` to save the entire working environment.


