--- 
title: "RBasic 101"
author: "Kyle Chung"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: everdark/rbasic
description: "The lecture notes for RBasic 101 on agilearn.io"
---

# Prologue

**The materials of this book target at new users or users not very familiar with R.** Before you start to read this book, take a look at the following examples to test your familiarity with the R language.

## Thought Experiments
### First Trial
```{r}
A <- 1 + 1
(B <- 1 + 1)
identical(A, B)
```

Now you know that adding parenthesis around an expression will force R to print the result of it. But why? If you immediately type `?"("` to search for a possible anwser, you are probably not new to the R language. And if you can further read the document and get yourself understood, you are not the targeted reader of this book.

### Second Trial
```{r}
a_vec <- 1:2
names(a_vec)
names(a_vec) <- c("foo", "bar")
names(a_vec)
```

The function `names` is used to return the names of the object given, if any. In the third line we also use `names` to rename the object. Now you know that we can use such tech to rename an object. But how is it possible *to asign a value to the returned value of a function*? 

Should you really think it is okay to do so, then think about the following codes:

```{r, error=TRUE}
xf <- function(x) x
xf(1) <- 2
```

Oops! Acutally you can't do that. Asign a value to the returned value of a function causes an error. But `names(a_vec) <- c("foo", "bar")` does work, right? If you do understand the reason why it actually works, you are not the targeted reader of this book. 

### Third Trial
```{r}
`+`(1, 2)
```

You may never see this coming before, or you just don't code in such style, but you do understand this is a function call to the add function, in a functional way. Then you are not the targeted reader of this book.

## Welcome to the R Language

If all the above examples bother you nothing, you are not the targeted reader of this book since you are quite familiar with the R language and you do know how to dive depper on your own. But if some or all of the above examples make you confused or starting googling, you can learn from this book.

This book is NOT a HOW-TO cookbook to teach you how to write in the R language. A HOW-TO cookbook will let you know, in the R language, how to force print an expression, or rename an object. This book is NOT a HOW-TO cookbook. Yes we can use `(` to force print an expression. But why? Yes we can use `names` to rename an object. 
But why? This book is to teach you WHY we write R in this or that way.  Consequently you are not just coding in R, you are also thinking in R. And that's where the R language starts to shine.






